Towards interactive model checking\\of biological pathway dynamics

### 2015-01-22 Paper_FORTE15.tex
Many many adjustments

A number of software tools are available
for modeling complex networks of biochemical interactions~\cite{bio-pepa,blenx,copasi,e-cell,gna}.
These tools significantly contribute to the process of formalizing the knowledge on biological
processes, rendering them amenable to computational analysis.
However, a lack of familiarity with the formalisms underlying many available tools
hampers their direct application by biology experts.
ANIMO (Analysis of Networks with Interactive MOdeling,~\cite{animo-site,animo-ieee,animo-gene}) is a software tool
based on the formalism of \tas~\cite{timed-automata-alur} that supports the
modeling of biological signaling pathways
by adding a dynamic component to traditional static representations of signaling networks.
ANIMO allows to compare the behavior of
a model with wet-lab data, and to explore such behavior in a user-friendly way.
In order to achieve a good level of user-friendliness for a public of biologists, the complexity of \tas\ is hidden \emph{under the hood},
presenting ANIMO as a plug-in to Cytoscape~\cite{cytoscape}, a tool specifically developed for visualizing 
and elaborating biological networks. Signaling networks are represented in ANIMO using
the nodes-edges notation normally used in the context of biology (see Fig.~\ref{fig:small-example-net}).

Currently, ANIMO supports interactive exploration of network dynamics based on simulation runs. 
Model checking queries can be answered through the UPPAAL tool~\cite{uppaal}, but
the required knowledge of temporal logic together with the usually long response times slows down the investigation process.
In order to pave the way for a widespread use of model checking on non-trivial models of
signaling networks, we propose here a new \tas\ model that marks a relevant improvement in terms of
performances with respect to the model currently used in ANIMO. Moreover, consistently with the intents of our tool, we
present also a user interface for the definition of model checking queries in ANIMO. This will allow a
user to interrogate an ANIMO model without previous experience in temporal logics.

@@

\begin{figure}[htbp]
\centering
\subfloat[\label{fig:small-example-net}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto_networkOnly}} \qquad
\subfloat[\label{fig:small-example01}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto01}} \qquad
\subfloat[\label{fig:small-example02}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto02}} \qquad
\subfloat[\label{fig:small-example03}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto03}}
\caption{A kinase signaling network represented in the nodes-edges form {\bf\protect\subref{fig:small-example-net}}
and its evolution represented by abstract molecular interactions
({\bf\protect\subref*{fig:small-example01}\--{}\protect\subref*{fig:small-example03}}).
{\sf T} is the target of the signaling pathway, {\sf TK}
is the kinase that activates {\sf T}, and {\sf TKK} is the kinase that activates {\sf TK}.
{\bf\protect\subref{fig:small-example01}} An already active (yellow) {\sf TKK} can bind to
an inactive (red) {\sf TK}, catalyzing its phosphorylation (i.e. binding of a phosphate group, {\sf P}).
This causes a change in the shape of {\sf TK}, activating its enzymatic function {\bf\protect\subref{fig:small-example02}}.
The active {\sf TK} can in turn activate its target {\sf T}, enabling it to carry out its function {\bf\protect\subref{fig:small-example03}}.
% {\bf\protect\subref*{fig:small-example-net}} represents the initial state of the network in the
% traditional nodes-edges form.
}\label{fig:small-example-biology}
\end{figure}

@@

A model in ANIMO is defined through the Cytoscape plug-in user interface (see Fig.~\ref{fig:cytoscape}), where the user inserts
a node for each molecular species and an edge for each reaction, with $\rightarrow$ indicating activation
and $\dashv$ indicating inhibition similarly to traditional representations of signaling networks (Fig.~\ref{fig:small-example-net}).

\begin{figure}[htb]
\begin{center}
   \includegraphics[width=.6\textwidth]{images/mapk_model_egf4}
\end{center}
\caption{The Cytoscape user interface running the ANIMO plug-in
(image from~\cite{animo-ieee}). The \emph{Network} panel in the center contains the nodes-edges
model of the cross-talking signaling pathways of NGF and EGF, with
colors indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first hour of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colors nodes in the \emph{Network} are based.
The series with the {\sf \_data} suffix is experimental
data from~\cite{egf-ngf}, considering a treatment with 100~ng/ml EGF.
\label{fig:cytoscape}}
\end{figure}

@@

\section{A new way of modeling signaling pathways with \tas}\label{sec:animo-new}
We propose here a novel model to represent signaling pathways with \tas\ in ANIMO.
We define the model previously used in ANIMO to be \emph{reaction-centered}, as for each reaction
in the network an instance of a \ta\ template is generated to mimic
the occurrences of that reaction. Observing that signaling
networks tend to be highly connected, containing noticeably more reactions than reactants,
we propose to shift the focus on reactants instead, achieving what we call a \emph{reactant-centered} model.
This change of view is inspired by the classical way in which biological events are modeled
with ordinary differential equations (ODEs)~\cite{ode-ma-anche-altro}.

@@

\subsection{The reactant-centered approach}\label{sec:reactant-centered}
The reactant-centered model presented here is based on the concept of \emph{net effect} of a set of reactions on a reactant:
instead of considering each reaction in isolation, we consider their combined influence on each reactant.
As an example, consider a reactant $A$ activated by reactions $R_1$ and $R_2$, and inhibited
by reaction $R_3$ (see Fig.~\ref{fig:network-abcd}).
The net effect of these three reactions on $A$ defines the \emph{net reaction} $R_A = R_1 + R_2 - R_3.$
Applying a concept similar to the definition of an ODE,
where the rate of change of each reactant depends on the rate of the reactions influencing it,
the rate of $R_A$ is computed as the sum of the rates of the reactions influencing~$A$: 
$$r_A = r_1 + r_2 - r_3$$
where $r_i$ is the rate of reaction $R_i$ and is defined as follows.
Consider $R_1$ to be the reaction $B \rightarrow A$ with kinetic constant $k_1$.
Suppose the settings in the ANIMO network for $R_1$ make its rate depend only on the
activity level of $B$. Then we compute the rate of $R_1$ as $r_1 = [B] \times k_1$, with
$[B]$ the current activity level of $B$.

@@

\subsection{Computations on the fly}\label{sec:rates-ta-model}
Reaction rates are all computed at run-time by the function {\sf update()},
and are based on the user-chosen scenarios, their kinetic constant $k$ and the activity level of the involved reactants.
As such computations require floating-point precision but UPPAAL only provides integer
variables and operators, we use a significand-and-exponent notation with 4 significant figures, which allows for an error
in the order of $0.1 \%$ while 
avoiding integer overflow in UPPAAL's engine.
For example, the floating point number $a = 1.23456$ will be represented as the pair $\langle 1235, -3 \rangle$,
which is translated back as $a = 1235 \times 10^{-3} = 1.235$.
% As an example of such a notation, consider the two floating-point numbers
% $a = 1.23456$ and $b = 0.00023456$, and their sum $c = a + b = 1.23479456$. The corresponding representations in
% our integer-based notation will be:
% $$
% \begin{array}{rcl}
%   a = \langle 1235, -3 \rangle &=& 1235 \times 10^{-3}\\
%   b = \langle 2346, -7 \rangle &=& 2346 \times 10^{-7}\\
%   a + b = \langle 1237, -3 \rangle &=& 1237 \times 10^{-3}
% \end{array}
% $$
% In this case, the number representing the sum $a + b$ differs from the exact result by $c - 1237 \times 10^{-3} = 0.001456$.
The interested reader can find the UPPAAL definitions and functions needed to compute rate and time values for the \tas\ templates
% in Appendix~\ref{apx:double-functions} or 
inside any UPPAAL model file generated by ANIMO with a reactant-centered model type~\footnote{For further instructions, see
the ANIMO manual at \url{http://fmt.cs.utwente.nl/tools/animo/manual.html}}.

@@

Being based on the same abstractions, the queries the two models can be analyzed with the same
UPPAAL queries for all the cases for which biology is involved. The differences can be seen only if the
analysis focuses on the single automaton level and explicitly refers to locations: such type of analysis
is normally not necessary when considering the model as a whole and using it for what it is, i.e. an
abstract representation of a set of biological processes.

We note that replacing all the existing reactions with the net reactions relative to each
reactant in a model does change the behavior of the model in the sense that continuous oscillations
of activity levels around an equilibrium point are less frequent. However, this does not mean that the
behavior of two models using the two different approaches will diverge: the equilibrium points
are still the same.
Two opposite reactions occurring with the same (or similar) frequency will (nearly) cancel each other out: the
behavior of the single reactions is explicitly represented in the reaction-centered model and gives rise to more oscillations.
On the other hand, in the reaction-centered model, the trajectory of a reactant's activity level
is determined taking into account also the cancelling effect of opposite reactions, so the simulations
will generally contain less oscillations.
From the user's point of view, graphs will look smoother while still exhibiting the same qualitative behavior,
as can be seen in Figure~\ref{fig:comparison-graph-animo}.
The \tas\ templates in Figure~\ref{fig:ta-models} show that the main difference between the two approaches
is indeed due to the choice of lumping all the reactions influencing $A$ into the single
reaction $R_A$.
% Exhibiting less oscillations means generating less events (i.e. changes of state) in the underlying \tas\ model, as in both approaches each change
% in reactant activity level is caused by an event in the \tas\ model. Less events means smaller state spaces, thus lower analysis times.
In the next session we will show that having less oscillations leads in practice to better analysis performances.
\begin{figure}[!htb]
\begin{center}
  \subfloat[Reaction-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reaction_10levels}\label{fig:comparison-graph-animo-reaction}}\qquad\quad
  \subfloat[Reactant-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reactant_10levels}\label{fig:comparison-graph-animo-reactant}}
\end{center}
  \caption{Comparison of two ERK activity graphs generated from the model based on the case study from~\cite{animo-ieee} with treatment of 50 ng/ml NGF.
  ERK was set to 10 levels of granularity to evidence the oscillations in the graphs.
  \label{fig:comparison-graph-animo}}
\end{figure}

@@

All experiments were carried out on an Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 14.04.1 64bit.
UPPAAL version 4.1.19 was used to compute the result of the queries, asking for ``some trace'' with random depth-first search order
when an execution trace was expected to be produced. For the simulation queries using the statistical model checking engine, we left
all options at their default values.

@@

\subsection{Simulation cost}
We start by evaluating the cost of simulation with the two different models. This is a particularly
important aspect to consider, as during the model building phase a user may need to perform a large number of simulations,
continuously adapting the topology or quantitative parameters of a network model. In order to make
the modeling approach in ANIMO as interactive as possible, it is desirable to decrease dead times,
and this translates into reducing the computational cost of model analysis as much as possible.
UPPAAL's statistical model checking engine~\cite{uppaal-smc} is based on the generation of random walks (i.e., simulation runs) from a \tas\ model.
In this experiment, we query UPPAAL for simulation runs on the models
generated by ANIMO applying the reaction- and reactant-centered modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the activity level of node NGF to 15/15, while changing EGF to be at 0/15 activity.
% We have chosen this configuration as the
% model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
% This makes the model more demanding when generating the state space
% because having more reactions occur means having more states in the transition system underlying the \tas\ model.
This configuration was chosen as it generates a more interesting behavior from the biological point of view,
also w.r.t. the model checking queries in Section~\ref{subsec:model-checking-performances};
the treatment with 100~ng/ml EGF was also tested and gave similar performance results.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the reactant-centered model, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network.

\begin{table}
  \begin{center}
  \begin{tabular}{|l||r|r|}
    \hline
    Model type & Time (s) & Memory (peak KB) \\
    \hline
    \hline
    Reaction-centered & 37.14 & 257,928 \\
    \hline
    Reactant-centered & 2.83 & 33,464 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the query {\tt simulate 100 [36000] \{ R1, R2, \dots, R11 \}} on the model from~\cite{animo-ieee} with starting condition
  NGF = 15/15, EGF = 0/15, corresponding to a treatment with 50 ng/ml NGF.
  The query asks for 100 time series of the activity levels of all reactants in the model over the first 60 minutes
  of execution.\label{tab:sim-100}}
\end{table}

@@

\begin{table*}[htbp]
  \begin{center}
    \begin{tabular}{|c||r|r||r|r||r|r|}
      \hline
       \multirow{3}{*}{Query} & \multicolumn{2}{c||}{Reaction-centered}	   & \multicolumn{2}{c||}{Reactant-centered} & \multicolumn{2}{c|}{Improvement}\\
      \cline{2-7}
       & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Memory} \\
       & \multicolumn{1}{c|}{time (s)}    & \multicolumn{1}{c||}{(peak KB)}    & \multicolumn{1}{c|}{time (s)} & \multicolumn{1}{c||}{(peak KB)} & \multicolumn{1}{c|}{(n-fold)} & \multicolumn{1}{c|}{(n-fold)}\\
      \hline
      \hline
      (1) & 127.25 & 351\,{}804 & 1.42 & 25\,{}716 & 90 & 14 \\
      \hline
      (2) & 167.64 & 269\,{}932 & 1.93 & 26\,{}112 & 87 & 10 \\
      \hline
      (3) & 150.90 & 318\,{}460 & 1.44 & 26\,{}136 & 105 & 12 \\
      \hline
      (4) & 2.15 & 216\,{}920 & 0.11 & 25\,{}316 & 19 & 9 \\
      \hline
      (5) & 470.56 & 311\,{}568 & 2.22 & 26\,{}648 & 212 & 12 \\
      \hline
      (6) & 453.90 & 357\,{}980 & 2.23 & 26\,{}652 & 204 & 13 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the given queries on the case study from~\cite{animo-ieee}.\label{tab:model-checking}}
\end{table*}

@@

\subsection{Analysis of the results}
% Note that we focus most of our attention on the NGF treatment, as with this configuration the model
% was hypothesized to keep reacting permanently (sustained ERK activity). The results from queries (1) and (2)
% show that this is indeed the case, as EGF treatment leads to all reactions dying out after a number of steps (i.e., deadlock due to all
% automata being in the {\sf not\_reacting} location).
% This means that the queries (3) - (6) deal with a larger state space, leading to presumably lower model checking performances.
Requesting a full inspection of the state space
as we do when using a query of the type {\tt A[]$\phi$} returning true in cases (5) and (6), allows us to indirectly compare the state space size of the two
model versions. As the hundreds-fold computation time improvements in Table~\ref{tab:model-checking} show, the reactant-centered model produces
indeed a noticeably smaller state space, allowing for a higher level of interactivity also when performing non-trivial model checking.
Moreover, our experiments point out that the reactant-centered approach considerably lowers the memory
requirements for the model. This is not only due to the absence of possibly large precomputed time tables,
which can contain thousands of elements each.
% \footnote{In the reaction-centered model, having a reaction depend on two 101-levels reactants
% leads to the generation of two time tables containing $101 \times 101 = 10\,{}201$ elements each.}
Indeed, this point was further investigated by implementing a reaction-centered model which
avoids the use of tables and instead makes on-the-fly computations of the time bounds with the same number representation as in the reactant-centered model.
This resulted in improved performances in the cases of reachability and simulation-based queries, with memory requirements similar to the ones for
the reactant-centered model. However, in all other cases a much larger amount of memory (more than 2 Gb) was used
with respect to the table-based implementation of the same model, without leading to appreciable benefits in terms of execution time:
in some cases performances were noticeably deteriorated.
% These findings support the intuitive idea that a reactant-centered approach
% leads in general to fewer events in a model, and to a smaller state space.
These findings seem to support the idea that a reactant-centered approach
leads in general to a smaller state space.

>>>>

A number of software tools are available
for modeling complex networks of biochemical interactions~\cite{bio-pepa,blenx,copasi,e-cell,gna}.
These tools significantly contribute to the process of formalizing the knowledge on biological
processes, rendering them amenable to computational analysis.
However, a lack of familiarity with the formalisms underlying many available tools
hampers their direct application by biology experts.
ANIMO (Analysis of Networks with Interactive MOdeling,~\cite{animo-site,animo-ieee,animo-gene}) is a software tool
based on the formalism of \tas~\cite{timed-automata-alur} that supports the
modeling of biological signaling pathways
by adding a dynamic component to traditional static representations of signaling networks.
ANIMO allows to compare the behavior of
a model with wet-lab data, and to explore such behavior in a user-friendly way.
In order to achieve a good level of user-friendliness for a public of biologists, the complexity of \tas\ is hidden \emph{under the hood},
presenting ANIMO as an app for Cytoscape~\cite{cytoscape}, a tool specifically developed for visualizing 
and elaborating biological networks. Signaling networks are represented in ANIMO using
the nodes-edges notation normally used in the context of biology (see Fig.~\ref{fig:small-example-net}).

Previously, ANIMO supported only interactive exploration of network dynamics based on simulation runs. 
Model checking queries could be answered through the UPPAAL tool~\cite{uppaal}, but
the required knowledge of temporal logic together with the usually long response times slowed down the investigation process.
In order to pave the way for a widespread use of model checking on non-trivial models of
signaling networks, we updated ANIMO with a new \tas\ model (presented here). This marks a relevant improvement in terms of
performances with respect to the model previously used in ANIMO. Moreover, consistently with the intents of our tool, we
implemented also a user interface for the definition of model checking queries in ANIMO. This allows a
user to interrogate an ANIMO model without requiring previous experience in temporal logics. These new
features are available in the latest version of ANIMO, which was recently reimplemented as a Cytoscape 3 app~\cite{animo-app-site}:
this lets users profit of the additional analysis features made available by the other apps in the Cytoscape environment.

@@

\begin{figure}[htbp]
\centering
\subfloat[\label{fig:small-example-net}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto_networkOnly}} \qquad \qquad
\subfloat[\label{fig:small-example01}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto01}} \qquad \qquad
\subfloat[\label{fig:small-example02}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto02}} \qquad \qquad
\subfloat[\label{fig:small-example03}]{\includegraphics[scale=\scalaRete]{images/tkk_tk_t_tutto03}}
\caption{A kinase signaling network represented in the nodes-edges form {\bf\protect\subref{fig:small-example-net}}
and its evolution represented by abstract molecular interactions
({\bf\protect\subref*{fig:small-example01}\--{}\protect\subref*{fig:small-example03}}).
{\sf T} is the target of the signaling pathway, {\sf TK}
is the kinase that activates {\sf T}, and {\sf TKK} is the kinase that activates {\sf TK}.
{\bf\protect\subref{fig:small-example01}} An already active (yellow) {\sf TKK} can bind to
an inactive (empty) {\sf TK}, catalyzing its phosphorylation (i.e. binding of a phosphate group, {\sf P}).
This causes a change in the shape of {\sf TK}, activating its enzymatic function {\bf\protect\subref{fig:small-example02}}.
The active {\sf TK} can in turn activate its target {\sf T}, enabling it to carry out its function {\bf\protect\subref{fig:small-example03}}.
% {\bf\protect\subref*{fig:small-example-net}} represents the initial state of the network in the
% traditional nodes-edges form.
}\label{fig:small-example-biology}
\end{figure}

@@

As ANIMO is a Cytoscape app, models are defined through the Cytoscape user interface (see Fig.~\ref{fig:cytoscape}), where the user inserts
a node for each molecular species and an edge for each reaction, with $\rightarrow$ indicating activation
and $\dashv$ indicating inhibition similarly to traditional representations of signaling networks (Fig.~\ref{fig:small-example-net}).

\begin{figure}[htb]
\begin{center}
   \includegraphics[width=.7\textwidth]{images/ANIMO3_MAPK_model} %mapk_model_egf4}
\end{center}
\caption{The Cytoscape 3 user interface running the new ANIMO app
(model from~\cite{animo-ieee}). The \emph{Network} panel in the center contains the nodes-edges
model of the cross-talking signaling pathways of NGF and EGF, with
colors indicating node activity levels and shapes representing different protein categories (see the \emph{Legend} on the left).
The \emph{Results Panel} on the right contains a graph plotting activity levels of selected nodes
during the first hour of evolution of the model. The slider under the graph
allows the user to select the time instant (marked as a vertical red line in the graph) on which
the colors nodes in the \emph{Network} are based. The edge thickness is used to give an idea on which reactions are
occurring more often at the selected time instant.
The series {\sf Erk (EGF) data} in the graph is the experimental
data from~\cite{egf-ngf} for the 100~ng/ml EGF treatment.
\label{fig:cytoscape}}
\end{figure}

@@

\section{A new way of modeling signaling pathways with \tas}\label{sec:animo-new}
We present here a novel model to represent signaling pathways with \tas\ in ANIMO.
We define the model previously used in ANIMO to be \emph{reaction-centered}, as for each reaction
in the network an instance of a \ta\ template is generated to mimic
the occurrences of that reaction. Observing that signaling
networks tend to be highly connected, containing noticeably more reactions than reactants,
we shift the focus on reactants instead, achieving what we call a \emph{reactant-centered} model.
This change of view is inspired by the classical way in which biological events are modeled
with ordinary differential equations (ODEs)~\cite{ode-ma-anche-altro}.

@@

\subsection{The reactant-centered approach}\label{sec:reactant-centered}
The reactant-centered model presented here is based on the concept of \emph{net effect} of a set of reactions on a reactant:
instead of considering each reaction in isolation, we consider their combined influence on each reactant.
As an example, consider a reactant $A$ activated by reactions $R_1$ and $R_2$, and inhibited
by reaction $R_3$ (see Fig.~\ref{fig:network-abcd}).
The net effect of these three reactions on $A$ defines the \emph{net reaction} $R_A = R_1 + R_2 - R_3.$
Applying a concept similar to the definition of an ODE,
where the rate of change of each reactant depends on the rate of the reactions influencing it,
the rate of $R_A$ is computed as the sum of the rates of the reactions influencing~$A$: 
$$r_A = r_1 + r_2 - r_3$$
where $r_i$ is the rate of reaction $R_i$ and is defined as follows.
Consider $R_1$ to be the reaction $B \rightarrow A$ with kinetic constant $k_1$.
Suppose the settings in the ANIMO network for $R_1$ make its rate depend only on the
activity level of $B$. Then we compute the rate of $R_1$ as $r_1 = [B] \times k_1$, with
$[B]$ the current activity level of~$B$.

@@

\subsection{Computations on the fly}\label{sec:rates-ta-model}
Reaction rates are all computed at run-time by the function {\sf update()},
and are based on the user-chosen scenarios, their kinetic constant $k$ and the activity level of the involved reactants.
As such computations require floating-point precision but UPPAAL only provides integer
variables and operators, we use a significand-and-exponent notation with 4 significant figures, which allows for an error
in the order of $0.1 \%$ while 
avoiding integer overflow in UPPAAL's engine.
For example, the floating point number $a = 1.23456$ will be represented as the pair $\langle 1235, -3 \rangle$,
which is translated back as $a = 1235 \times 10^{-3} = 1.235$.
% As an example of such a notation, consider the two floating-point numbers
% $a = 1.23456$ and $b = 0.00023456$, and their sum $c = a + b = 1.23479456$. The corresponding representations in
% our integer-based notation will be:
% $$
% \begin{array}{rcl}
%   a = \langle 1235, -3 \rangle &=& 1235 \times 10^{-3}\\
%   b = \langle 2346, -7 \rangle &=& 2346 \times 10^{-7}\\
%   a + b = \langle 1237, -3 \rangle &=& 1237 \times 10^{-3}
% \end{array}
% $$
% In this case, the number representing the sum $a + b$ differs from the exact result by $c - 1237 \times 10^{-3} = 0.001456$.
The interested reader can find the UPPAAL definitions and functions needed to compute rate and time values for the \tas\ templates
% in Appendix~\ref{apx:double-functions} or 
inside any UPPAAL model file generated by ANIMO with a reactant-centered model type~\footnote{Models generated
by ANIMO are saved in the system's temporary directory. Further details are available in
the ANIMO manual at \url{http://fmt.cs.utwente.nl/tools/animo/content/Manual.pdf}}.

@@

Being based on the same abstractions, the two models can be analyzed with the same
UPPAAL queries for all the cases for which biology is involved. The differences can be seen only if the
analysis focuses on the single automaton level and explicitly refers to locations: such type of analysis
is normally not necessary when considering the model as a whole and using it for what it is, i.e. an
abstract representation of a set of biological processes.

We note that replacing all the existing reactions with the net reactions relative to each
reactant in a model does change the behavior of the model in the sense that continuous oscillations
of activity levels around an equilibrium point are less frequent. However, this does not mean that the
behavior of two models using the two different approaches will diverge: the equilibrium points
are still the same.
Two opposite reactions occurring with the same (or similar) frequency will (nearly) cancel each other out: the
behavior of the single reactions is explicitly represented in the reaction-centered model and gives rise to more oscillations.
On the other hand, in the reactant-centered model, the trajectory of a reactant's activity level
is determined taking into account also the cancelling effect of opposite reactions, so the simulations
will generally contain less oscillations.
From the user's point of view, graphs will look smoother while still exhibiting the same qualitative behavior,
as can be seen in Figure~\ref{fig:comparison-graph-animo}.
The \tas\ templates in Figure~\ref{fig:ta-models} show that the main difference between the two approaches
is indeed due to the choice of lumping all the reactions influencing $A$ into the single
reaction $R_A$.
% Exhibiting less oscillations means generating less events (i.e. changes of state) in the underlying \tas\ model, as in both approaches each change
% in reactant activity level is caused by an event in the \tas\ model. Less events means smaller state spaces, thus lower analysis times.
In the next session we will show that having less oscillations leads in practice to better analysis performances.
\begin{figure}[!htb]
\begin{center}
  \subfloat[Reaction-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reaction_10levels}\label{fig:comparison-graph-animo-reaction}}\qquad\qquad
  \subfloat[Reactant-centered]{\includegraphics[scale=0.07]{images/grafico_ngf_reactant_10levels}\label{fig:comparison-graph-animo-reactant}}
\end{center}
  \caption{Comparison of two ERK activity graphs generated from the model based on the case study from~\cite{animo-ieee} with treatment of 50 ng/ml NGF.
  ERK was set to 10 levels of granularity to evidence the oscillations in the graphs.
  \label{fig:comparison-graph-animo}}
\end{figure}

@@

All experiments were carried out on an Intel\circledR\ Core$^{\mbox{\scriptsize\texttrademark}}$ i7 CPU at 2.80GHz equipped with 4 Gb RAM
and running Ubuntu GNU/Linux 14.04.1 64bit.
UPPAAL version 4.1.19 64bit was used to compute the result of the queries, asking for ``some trace'' with random depth-first search order
when an execution trace was expected to be produced. For the simulation queries using the statistical model checking engine, we left
all options at their default values.

@@

\subsection{Simulation cost}
We start by evaluating the cost of simulation with the two different models. This is a particularly
important aspect to consider, as during the model building phase a user may need to perform a large number of simulations,
continuously adapting the topology or quantitative parameters of a network model. In order to make
the modeling approach in ANIMO as interactive as possible, it is desirable to decrease dead times,
and this translates into reducing the computational cost of model analysis as much as possible.
UPPAAL's statistical model checking engine~\cite{uppaal-smc} is based on the generation of random walks (i.e., simulation runs) from a \tas\ model.
In this experiment, we query UPPAAL for simulation runs on the models
generated by ANIMO applying the reaction- and reactant-centered modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the activity level of node NGF to 15/15, while changing EGF to be at 0/15 activity.
% We have chosen this configuration as the
% model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
% This makes the model more demanding when generating the state space
% because having more reactions occur means having more states in the transition system underlying the \tas\ model.
This configuration was chosen as it generates a more interesting behavior from the biological point of view,
also w.r.t. the model checking queries in Section~\ref{subsec:model-checking-performances};
the treatment with 100~ng/ml EGF was also tested and gave similar performance results.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $91 \%$ less time with the reactant-centered model, using $97 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network.

\begin{table}
  \begin{center}
  \begin{tabular}{|l||r|r|}
    \hline
    Model type & Time (s) & Memory (peak KB) \\
    \hline
    \hline
    Reaction-centered & 30.72 & 291\,{}576 \\ %37.14 & 257,928 \\
    \hline
    Reactant-centered & 2.86 & 9\,{}768 \\ %2.83 & 33,464 \\
    \hline
  \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the query {\tt simulate 100 [36000] \{ R1, R2, \dots, R11 \}} on the model from~\cite{animo-ieee} with starting condition
  NGF = 15/15, EGF = 0/15, corresponding to a treatment with 50 ng/ml NGF.
  The query asks for 100 time series of the activity levels of all reactants in the model over the first 60 minutes
  of execution.\label{tab:sim-100}}
\end{table}

@@

\begin{table*}[htbp]
  \begin{center}
    \begin{tabular}{|c||r|r||r|r||r|r|}
      \hline
       \multirow{3}{*}{Query} & \multicolumn{2}{c||}{Reaction-centered}	   & \multicolumn{2}{c||}{Reactant-centered} & \multicolumn{2}{c|}{Improvement}\\
      \cline{2-7}
       & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Computation} & \multicolumn{1}{c||}{Memory usage} & \multicolumn{1}{c|}{Time} & \multicolumn{1}{c|}{Memory} \\
       & \multicolumn{1}{c|}{time (s)}    & \multicolumn{1}{c||}{(peak KB)}    & \multicolumn{1}{c|}{time (s)} & \multicolumn{1}{c||}{(peak KB)} & \multicolumn{1}{c|}{(n-fold)} & \multicolumn{1}{c|}{(n-fold)}\\
      \hline
      \hline
      (1) & 126.56 & 523\,{}448 & 1.04 & 9\,{}236 & 122 & 57 \\ %127.25 & 351\,{}804 & 1.42 & 25\,{}716 & 90 & 14 \\
      \hline
      (2) & 159.29 & 436\,{}496 & 1.73 & 11\,{}480 & 92 & 38 \\ %167.64 & 269\,{}932 & 1.93 & 26\,{}112 & 87 & 10 \\
      \hline
      (3) & 146.09 & 439\,{}484 & 1.04 & 10\,{}384 & 140 & 42 \\ %150.90 & 318\,{}460 & 1.44 & 26\,{}136 & 105 & 12 \\
      \hline
      (4) & 0.74 & 293\,{}508 & 0.06 & 7\,{}484 & 12 & 39 \\ %2.15 & 216\,{}920 & 0.11 & 25\,{}316 & 19 & 9 \\
      \hline
      (5) & 581.79 & 448\,{}764 & 6.86 & 16\,{}880 & 85 & 27 \\ %470.56 & 311\,{}568 & 2.22 & 26\,{}648 & 212 & 12 \\
      \hline
      (6) & 561.01 & 449\,{}248 & 6.42 & 15\,{}852 & 87 & 28 \\ %453.90 & 357\,{}980 & 2.23 & 26\,{}652 & 204 & 13 \\
      \hline
    \end{tabular}
  \end{center}
  \caption{UPPAAL processor time and memory usage for reaction- and reactant-centered modeling approaches when computing
  the given queries on the case study from~\cite{animo-ieee}.\label{tab:model-checking}}
\end{table*}

@@

\subsection{Analysis of the results}
% Note that we focus most of our attention on the NGF treatment, as with this configuration the model
% was hypothesized to keep reacting permanently (sustained ERK activity). The results from queries (1) and (2)
% show that this is indeed the case, as EGF treatment leads to all reactions dying out after a number of steps (i.e., deadlock due to all
% automata being in the {\sf not\_reacting} location).
% This means that the queries (3) - (6) deal with a larger state space, leading to presumably lower model checking performances.
Requesting a full inspection of the state space
as we do when using a query of the type {\tt A[]$\phi$} returning true in cases (5) and (6), allows us to indirectly compare the state space size of the two
model versions. As the computation time improvements in Table~\ref{tab:model-checking} show, the reactant-centered model produces
indeed a noticeably smaller state space, allowing for a higher level of interactivity also when performing non-trivial model checking.
Moreover, our experiments point out that the reactant-centered approach considerably lowers the memory
requirements for the model. This is not only due to the absence of possibly large precomputed time tables,
which can contain thousands of elements each.
% \footnote{In the reaction-centered model, having a reaction depend on two 101-levels reactants
% leads to the generation of two time tables containing $101 \times 101 = 10\,{}201$ elements each.}
Indeed, this point was further investigated by implementing a reaction-centered model which
avoids the use of tables and instead makes on-the-fly computations of the time bounds with the same number representation as in the reactant-centered model.
This resulted in improved performances in the cases of reachability and simulation-based queries, with memory requirements closer to the ones for
the reactant-centered model (0.69 s and 24\,{}796 KB for query (4)). However, in all other cases a much larger amount of memory (around 2 Gb) was used
with respect to the table-based implementation of the same model, without leading to appreciable benefits in terms of execution time:
in some cases performances were noticeably deteriorated (600-800 s for queries (1)-(3)).
% These findings support the intuitive idea that a reactant-centered approach
% leads in general to fewer events in a model, and to a smaller state space.
These findings seem to support the idea that a reactant-centered models
have a smaller state space.
